---
title: 'Data-driven evidence that amyloid accumulation in the brain alters lifespan trajectory of aging'
author: "Brian B. Avants"
date: "August 17, 2016"
output: html_document
---

The Dallas Lifespan Brain Study released an imaging/cognition/AV45 dataset publicly consisting of 315 subjects.  Of these subjects, 127 (FIXME) have amyloid imaging.

```{r}
library( ANTsR )
bd = path.expand( "~/Downloads/DLBS/")
template = antsImageRead( "~/data/templates/OasisTemplateTustisonEtAl/T_template0_BrainCerebellum.nii.gz")
templateSeg = antsImageRead( "~/data/templates/OasisTemplateTustisonEtAl/T_template0_glm_6labelsJointFusion.nii.gz")
```

Get the demographics and merge with cognition.

```{r, echo=FALSE}
demog = read.csv( paste( bd, "cognitive_data/Subject_Information.csv",sep=''))
n=ncol(demog)
mycsvs = rev(Sys.glob( paste( bd, "cognitive_data/*sv",sep='' ) ))[-c(1)]
for ( k in mycsvs ) {
  temp = read.csv( k )
  print( dim( temp ) )
  print( sum( is.na( temp[,] ) ) )
  demog = merge( demog, temp, by = "INDI_ID" )
  }
```

Find all AV45, jacobian and thickness images.

```{r fusemodalities}
dirs = paste( bd, c("pet/", "thickness/", "jacobian/", 'segmentations/'), sep='' )
exts = c( "AV45SUVR.nii.gz", 
          "CorticalThicknessNormalizedToTemplate.nii.gz", 
          "SubjectToTemplateLogJacobian.nii.gz",
          "BrainSegmentation.nii.gz" )
fnmat = matrix( NA, nrow=nrow( demog ), ncol = length( dirs ) )
for ( i in 1:nrow( demog ) )
  {
  mypre = paste( dirs, demog$INDI_ID[i], "*", exts, sep='' )
  if ( length( Sys.glob( mypre[1] ) ) > 0 ) fnmat[i,1] = Sys.glob( mypre[1] )
  if ( length( Sys.glob( mypre[2] ) ) > 0 ) fnmat[i,2] = Sys.glob( mypre[2] )
  if ( length( Sys.glob( mypre[3] ) ) > 0 ) fnmat[i,3] = Sys.glob( mypre[3] )
  if ( length( Sys.glob( mypre[3] ) ) > 0 ) fnmat[i,4] = Sys.glob( mypre[4] )
  }
haveAmy = !is.na( fnmat[,1] )
```



collect segmentation volumes

```{r getMatsGlobalCor2}
if ( ! exists( "segvols" ) )
  {
  segvols = matrix( nrow=sum(haveAmy), ncol=6 )
  ct = 1
  for ( i in which( haveAmy ) )
    {
    seg = antsImageRead( fnmat[i,4] )
    volelt = sqrt( sum( antsGetSpacing( seg )^2 ) )
    for ( k in 1:6 ) segvols[ ct, k ] = volelt * sum( seg == k )
    ct = ct + 1
    }
  segvols = data.frame( segvols )
  colnames( segvols ) = c("CSF","GM","WM","DGM","BrainStem","Cerebellum")
  }
```

Convert scalar images to matrices.

```{r getMats}
if ( ! exists("amat") )
{
res = rep( 2, 3 ) # resolution at which to do study
smth = rep( 1, 3 )
cmask = thresholdImage( templateSeg, 2, 2 ) %>% resampleImage( res )
wmask = thresholdImage( templateSeg, 3, 3 ) %>% resampleImage( res )
dmask = thresholdImage( templateSeg, 4, 4 ) %>% resampleImage( res )
allmask = thresholdImage( templateSeg, 1, 6 ) %>% resampleImage( res )
mask14 = thresholdImage( templateSeg, 1, 4 ) %>% resampleImage( res )
alist = imageFileNames2ImageList( fnmat[haveAmy,1] )
amat = imageListToMatrix( alist, mask14, sigma = smth  )
tlist = imageFileNames2ImageList( fnmat[haveAmy,2] )
tmat = imageListToMatrix( tlist, cmask, sigma = smth )
jlist = imageFileNames2ImageList( fnmat[haveAmy,3] )
jmat = imageListToMatrix( jlist, allmask )
rm( jlist, tlist, alist )
for ( i in 1:10 ) gc()
# jwmat = imagesToMatrix( fnmat[haveAmy,3], wmask )
# jdmat = imagesToMatrix( fnmat[haveAmy,3], dmask )
}
```


how many clusters in the amyloid data?
```{r pamamy,eval=TRUE}
library( ggplot2 )
amatRMeans = rowMeans( amat )
hist( amatRMeans )
k=80
if ( ! exists( "scanner" )  ) {
  lowdecom = sparseDecom2(
    inmatrix = list( scale(amat), scale(jmat) ),
    inmask   = c( mask14, allmask ), verbose=F,
    sparseness = c( 0.1, 0.1), smooth = 0.5,
    mycoption = 0, perms=0,
    nvecs = 4, its = 15, cthresh = c( 10, 10 ) )
  names(lowdecom)[3]="v"
  }
# seconds
# lowdecom = rsvd::rsvd( (amat), k, nu=0, p=10, sdist='normal' )
# spmat  = sparseDistanceMatrix( amat,  k = 25, kmetric = 'cor' )
# lowdecom =  irlba::irlba( spmat , nv=100, nu=0, center=colMeans(spmat) , right_only=TRUE)
tempimg = abs( makeImage( mask14, lowdecom$v[,3] ) ) %>%
  iMath("Normalize")
plot( template  %>% resampleImage( res ), tempimg, nslices=20,
      ncolumns=10, axis=2, window.overlay=c(0.1,1) )
#
# lowdecom = irlba::irlba( amat, k, nu=0 ) # minutes
# plot( ts( lowdecom$d[2:k] ) )
krange = 1:6
lowmat = amat %*% ( lowdecom$v )
# lowmat = cbind( amat %*% ( scanner$eig1[,] ) )
#                tmat %*% ( scanner$eig2[,1:4] ) )
pk <- fpc::pamk( lowmat, krange = krange)
hcl = hclust( dist( lowmat ) )
myclusts = cutree( hcl, k = 2 )
knitr::kable( data.frame( table(myclusts) ) )
myclusts = pk$pamobject$clustering
knitr::kable( data.frame( table(myclusts) ) )
lowAmy <- data.frame( amy = amatRMeans[myclusts==1] )
lowAmy$status <- 'lowAmy'
amyStatus <- rbind( lowAmy )
uclusts = unique( myclusts )[-1]
print( paste( 1, mean(amatRMeans[myclusts==1] ) ))
for ( k in uclusts )
  {
  highAmy <- data.frame( amy = amatRMeans[myclusts==k]  )
  highAmy$status <- paste('Amy',k,sep='')
  amyStatus = rbind( amyStatus, highAmy )
  print( paste( k, mean(amatRMeans[myclusts==k] ) ))
  }
ggplot( amyStatus, aes(amy, fill = status)) + geom_density(alpha = 0.2)
amat2 = amat %*% lowdecom$v
```



correlate the means of these matrices and look at a simple regression

```{r getMatsGlobalCor}
temp = cbind( 
  rowMeans( amat ),
  rowMeans( tmat ) )
#  rowMeans( jwmat ),
#  rowMeans( jdmat ),
#  rowMeans( jmat ) )
colnames( temp ) = c( "Amyloid", "Thickness" ) # "WJac", "DGMJac", 
inds = c(2,3,4,9,11:ncol(demog)) # base and cognition
BV = rowSums( segvols )
locdemog = data.frame( demog[haveAmy, inds], temp,  BV=BV, 
  psych::winsor(segvols,0.01), clust=factor( myclusts ) )
# for ( k in unique( myclusts ) )
  {
  sel = myclusts != 0
  locdemogSub1 = locdemog[ sel,]
  formchar="~ Age + Thickness + factor(Gender) +  MMSE + 
#              RAVENS_Number.correct_1 +
#              HOP_immediate.recall_1 +
#              HOP_delayed.recognition.acc_1 +
              HOP_delayed.recall_1 + 
              DC_total_1 +
              CANTAB_SST_SSRT_last_half_1 + BV +
#              LNS_Total.Correct_1 +
    WM + GM " # + BrainStem + Cerebellum" 
  myform = as.formula(paste("Amyloid", formchar ))
  mdl1 = glm( myform, data = locdemogSub1, family='gaussian' )
   #GM + CSF + WM + BrainStem + Cerebellum, data = locdemogSub1 )
  print( summary( mdl1 ) )
  visreg::visreg( mdl1, "GM" )
  visreg::visreg( mdl1, "WM" )
  }
```



do a sparse decom of the tmat

```{r tmatdecom,eval=FALSE}
# sel = locdemog$Age > 58
tspmat  = sparseDistanceMatrix( tmat[,],  k = 25, kmetric = 'cor' )
lowdecom =  irlba::irlba( tspmat , nv=200, nu=0, 
  center=colMeans(tspmat) , right_only=TRUE)
tempimg = abs( makeImage( mask14, lowdecom$v[,6] ) ) %>%
  iMath("Normalize")
plot( template  %>% resampleImage( res ), tempimg, nslices=20,
      ncolumns=10, axis=2, window.overlay=c(0.21,1) )
# amat2 = amat %*% lowdecom$v
tmat2 = tmat %*% lowdecom$v
```

Do a higher resolution map of local amyloid.

```{r amyMatRegression}
formchar="  Age + factor(Gender) +
            MMSE +
            RAVENS_Number.correct_1 +
            HOP_delayed.recognition.acc_1 +
            HOP_delayed.recall_1 + 
            DC_total_1 + WM  + Thickness + BrainStem "
myform = as.formula(paste("amat2~", formchar ))
mdl = lm( myform, data = locdemog )
bmdl = bigLMStats( mdl )
for ( nm in rownames( bmdl$beta.pval ) ) {
  mnbh = min( p.adjust( bmdl$beta.pval[nm,], 'BH' ) )
  print( paste( nm, min( p.adjust( bmdl$beta.pval[nm,], 'none' ) ), mnbh ) )
  }
# wjac is strong ... let's look
nm = "Thickness"
wjtstat = makeImage( mask14,  bmdl$beta.t[nm,] )
```

Render the thing on the thing.

```{r renstat}
ap="~/code/ants-src/bin/bin/"
rp =            matrix( c(90,0,180), ncol = 3 )
rp = rbind( rp, matrix( c(90,180,180), ncol = 3 ) )
rp = rbind( rp, matrix( c(270,270,270), ncol = 3 ) ) # top view
rp = rbind( rp, matrix( c(270,90,90), ncol = 3 ) ) # keeper, bottom view
rng = c( 3, max( abs( wjtstat ) ) )
volfn = path.expand( "~/Downloads/temp2" )
antsrVol( template %>% resampleImage(res), list( abs(wjtstat) ), overlayLimits = rng, 
          intensityTruncation = c(0.2,0.8), rotationParams = rp,
          magnificationFactor = 2.5,
          antspath = ap, filename = volfn )
volfn = paste( volfn, '.png', sep='' )
```

Add the figure we just made.

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics( volfn )
```

Plot the stat.

```{r jMatRegressionViz0}
rng = c(3,max(abs(wjtstat)))
plot( template  %>% resampleImage( res ), wjtstat, nslices=20,
      ncolumns=10, axis=2, window.overlay=rng )
plot( template  %>% resampleImage( res ), wjtstat*(-1), nslices=20,
      ncolumns=10, axis=2, window.overlay=rng )
```

Define a subset of the population for which there is no difference in age 
and for which there is no difference in MMSE.

```{r agematch}
locsel = locdemog$Age >= 50
myclustsLevs = unique( myclusts )
for ( k in myclustsLevs )
  {
  ksel = myclusts == k
  print( paste("Cluster",k,"MeanAmy",mean(locdemog$Amyloid[ksel])))
  print( range(locdemog$Age[ksel] ))
  print( paste(
    t.test( locdemog[locsel & locdemog$clust==1,]$Age, 
          locdemog[locsel & locdemog$clust==k,]$Age )$p.value,
    t.test( locdemog[locsel & locdemog$clust==1,]$MMSE, 
          locdemog[locsel & locdemog$clust==k,]$MMSE )$p.value,
    t.test( locdemog[locsel & locdemog$clust==1,]$Amyloid, 
          locdemog[locsel & locdemog$clust==k,]$Amyloid )$p.value ))
  }
```

Do a higher resolution map of jacobian.

```{r jMatRegression,eval=TRUE}
formchar="  MMSE +
#            RAVENS_Number.correct_1 +
#            HOP_delayed.recognition.acc_1 +
#            HOP_delayed.recall_1 + 
#            DC_total_1 
            BrainStem + Cerebellum + 
  Variate000 +Variate001  "
tnm = "clust"
locsel = locdemog$Age >= 0
myform = paste( "",tnm," + Age + factor(Gender)+",formchar )
myform2 = as.formula(paste("tmat[locsel,]~", myform ))
locdemog2 = cbind( locdemog, amat2 )
mdl = lm( myform2, data = locdemog2[locsel , ] )
bmdl = bigLMStats( mdl )
for ( nm in rownames( bmdl$beta.pval ) ) {
  mnbh = min( p.adjust( bmdl$beta.pval[nm,], 'holm' ) )
  print( paste( nm, 
    min( p.adjust( bmdl$beta.pval[nm,], 'none' ) ), mnbh ) )
  }
# wjac is strong ... let's look
tnm=paste(tnm,"2",sep='')
# tnm="Age"
if ( sum( cmask ) == length(bmdl$beta.t[tnm,]) )
  amyjtstat = makeImage( cmask, bmdl$beta.t[tnm,] )
if ( sum( allmask ) == length(bmdl$beta.t[tnm,]) )
  amyjtstat = makeImage( allmask, bmdl$beta.t[tnm,] )
if ( sum( mask14 ) == length(bmdl$beta.t[tnm,]) )
  amyjtstat = makeImage( mask14, bmdl$beta.t[tnm,] )
antsImageWrite( amyjtstat, "~/Downloads/temp.nii.gz")
print( range( amyjtstat ) )
```


Render the thing on the thing.

```{r renstat0,eval=TRUE}
rng = c( 4, max( abs( amyjtstat ) ) )
antsrVol( template %>% resampleImage( res ), 
          list( abs(amyjtstat) ), overlayLimits = rng, 
          intensityTruncation = c(0.2,0.8), rotationParams = rp,
          magnificationFactor = 2.5,
          antspath = ap, filename = "~/Downloads/tempT" )
```

slice plot 

```{r jMatRegressionViz,eval=TRUE}
rng = c( 3, max( abs( amyjtstat ) ) )
plot( template %>% resampleImage( res ), abs(amyjtstat), 
 nslices=20, ncolumns=10, axis=2, window.overlay=rng )
```

Plot the max stat.

```{r maxjstat,eval=TRUE}
sel = bmdl$beta.pval[nm,] < 1.e-4
print( sum( sel ) )
jvox = rowMeans( tmat[, sel ] )
vdemog = data.frame( voxel = jvox, locdemog2 )
myform2 = as.formula(paste("voxel~ ", myform ))
mdl = lm( myform2, data = vdemog )
visreg::visreg( mdl, tnm  )
```


